---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---

## Title: Your document should have a title that briefly summarizes your data analysis

## Synopsis: Blah-Blah-Blah


## Data Processing
##### As part of data processing, we want to derive summarized values by Event Type for following-
###### Total Fatalities
###### Total Injuries
###### Total Fatalities and Injuries Combined
###### Total Dollar value of Property Damage
###### Total Dollar value of Crop Damage
###### Total Dollar value of Property Damage and Crop Damage Combined

##### Data Processing Steps-
###### 1. Read the csv file using read.csv function.
###### 2. Handle exponent value of PROPDMGEXP and CROPDMGEXP columns.
###### 3. Derive a multipler and store it into a new column in Storm dataset.
###### 4. Use this multiplier to derive total dollar vale of property damage and crop damage. Store these values in 2 separate columns in Storm dataset.
###### 5. Derive total dollar value of property damage and crop damage combined together.
###### 6. Derive total combined value of fatalities and injuries by adding up "FATALITIES" and "INJURIES". Store this value in a new column in Storm dataset.


```{r}
library(dplyr)
library(ggplot2)
setwd(
"/Users/abhishekdubey/abhishek-git-repos/datasciencecoursera/reproducible-research/course_project_2"
)
stormData <- read.csv(file = "repdata_data_StormData.csv", header = TRUE, stringsAsFactors = FALSE)
```
**Handle exponent value of PROPDMGEXP and CROPDMGEXP columns-<br>**
** These are possible values of CROPDMGEXP and PROPDMGEXP- H,h,K,k,M,m,B,b,+,-,?,0,1,2,3,4,5,6,7,8, and blank-character <br>**
** H,h = hundreds = 100 <br>**
** K,k = kilos = thousands = 1,000 <br>**
** M,m = millions = 1,000,000 <br>**
** B,b = billions = 1,000,000,000 <br>**
** (+) = 1 <br>**
** (-) = 0 <br>**
** (?) = 0 <br>**
** black/empty character = 0 <br>**
** numeric 0..8 = 10 <br>**
**I have defined a new column PROPDMG_CROPDMG_MULTIPLIER in the storm dataset. The value stored in this column is a numeric multiplier based on above mentioned conditions. We will multiply columns PROPDMG and CROPDMG with PROPDMG_CROPDMG_MULTIPLIER to derive the dollar value of "Property Damage" and "Crop Damage" respectively**

```{r}
stormData[stormData$PROPDMGEXP %in% c("-","?",""), "PROPDMG_CROPDMG_MULTIPLIER"] = 0
stormData[stormData$PROPDMGEXP %in% c("+"), "PROPDMG_CROPDMG_MULTIPLIER"] = 1
stormData[stormData$PROPDMGEXP %in% c("0","1","2","3","4","5","6","7","8"), "PROPDMG_CROPDMG_MULTIPLIER"] = 10
stormData[stormData$PROPDMGEXP %in% c("H","h"), "PROPDMG_CROPDMG_MULTIPLIER"] = 100
stormData[stormData$PROPDMGEXP %in% c("K","k"), "PROPDMG_CROPDMG_MULTIPLIER"] = 1000
stormData[stormData$PROPDMGEXP %in% c("M","m"), "PROPDMG_CROPDMG_MULTIPLIER"] = 1000000
stormData[stormData$PROPDMGEXP %in% c("B","b"), "PROPDMG_CROPDMG_MULTIPLIER"] = 1000000000

stormDataSummarized <-  select(stormData, 
                               EVTYPE, 
                               FATALITIES, 
                               INJURIES, 
                               PROPDMG, 
                               CROPDMG, 
                               PROPDMG_CROPDMG_MULTIPLIER) %>% 
                        mutate(FATALITIES_AND_INJURIES_COMBINED = FATALITIES + INJURIES) %>%
                        mutate(PROPDMG_DOLLAR_VALUE = PROPDMG*PROPDMG_CROPDMG_MULTIPLIER) %>%
                        mutate(CROPDMG_DOLLAR_VALUE = CROPDMG*PROPDMG_CROPDMG_MULTIPLIER) %>%
                        mutate(DOLLAR_VALUE_FOR_PROP_AND_CROP_DAMAGE_COMBINED = 
                                   PROPDMG_DOLLAR_VALUE + CROPDMG_DOLLAR_VALUE) %>%
                        select(EVTYPE, 
                               FATALITIES, 
                               INJURIES, 
                               FATALITIES_AND_INJURIES_COMBINED, 
                               PROPDMG_DOLLAR_VALUE, 
                               CROPDMG_DOLLAR_VALUE, 
                               DOLLAR_VALUE_FOR_PROP_AND_CROP_DAMAGE_COMBINED) %>%
                        group_by(EVTYPE) %>% 
                        summarize(TOTAL_FATALITIES = sum(FATALITIES),
                                  TOTAL_INJURIES = sum(INJURIES),
                                  TOTAL_FATALITIES_AND_INJURIES = 
                                      sum(FATALITIES_AND_INJURIES_COMBINED),
                                  TOTAL_PROPDMG_DOLLAR_VALUE = sum(PROPDMG_DOLLAR_VALUE),
                                  TOTAL_CROPDMG_DOLLAR_VALUE = sum(CROPDMG_DOLLAR_VALUE),
                                  TOTAL_DOLLAR_VALUE_FOR_PROP_AND_CROP_DAMAGE = 
                                      sum(DOLLAR_VALUE_FOR_PROP_AND_CROP_DAMAGE_COMBINED)
                                  ) %>%
                        select(
                                EVTYPE, 
                                TOTAL_FATALITIES, 
                                TOTAL_INJURIES, 
                                TOTAL_FATALITIES_AND_INJURIES, 
                                TOTAL_PROPDMG_DOLLAR_VALUE, 
                                TOTAL_CROPDMG_DOLLAR_VALUE, 
                                TOTAL_DOLLAR_VALUE_FOR_PROP_AND_CROP_DAMAGE
                               )
```

## Results

