---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---

## Title: Your document should have a title that briefly summarizes your data analysis

## Synopsis: Blah-Blah-Blah


## Data Processing

```{r}
library(dplyr)
library(ggplot2)
setwd(
"/Users/abhishekdubey/abhishek-git-repos/datasciencecoursera/reproducible-research/course_project_2"
)
stormData <- read.csv(file = "repdata_data_StormData.csv", header = TRUE, stringsAsFactors = FALSE)
```
**Handle exponent value of PROPDMGEXP and CROPDMGEXP columns-<br>**
** These are possible values of CROPDMGEXP and PROPDMGEXP- H,h,K,k,M,m,B,b,+,-,?,0,1,2,3,4,5,6,7,8, and blank-character <br>**
** H,h = hundreds = 100 <br>**
** K,k = kilos = thousands = 1,000 <br>**
** M,m = millions = 1,000,000 <br>**
** B,b = billions = 1,000,000,000 <br>**
** (+) = 1 <br>**
** (-) = 0 <br>**
** (?) = 0 <br>**
** black/empty character = 0 <br>**
** numeric 0..8 = 10 <br>**
**I have defined a new column PROPDMG_CROPDMG_MULTIPLIER in the storm dataset. The value stored in this column is a numeric multiplier based on above mentioned conditions. We will multiply columns PROPDMG and CROPDMG with PROPDMG_CROPDMG_MULTIPLIER to derive the dollar value of "Property Damage" and "Crop Damage" respectively**

```{r}
stormData[stormData$PROPDMGEXP %in% c("-","?",""), "PROPDMG_CROPDMG_MULTIPLIER"] = 0
stormData[stormData$PROPDMGEXP %in% c("+"), "PROPDMG_CROPDMG_MULTIPLIER"] = 1
stormData[stormData$PROPDMGEXP %in% c("0","1","2","3","4","5","6","7","8"), "PROPDMG_CROPDMG_MULTIPLIER"] = 10
stormData[stormData$PROPDMGEXP %in% c("H","h"), "PROPDMG_CROPDMG_MULTIPLIER"] = 100
stormData[stormData$PROPDMGEXP %in% c("K","k"), "PROPDMG_CROPDMG_MULTIPLIER"] = 1000
stormData[stormData$PROPDMGEXP %in% c("M","m"), "PROPDMG_CROPDMG_MULTIPLIER"] = 1000000
stormData[stormData$PROPDMGEXP %in% c("B","b"), "PROPDMG_CROPDMG_MULTIPLIER"] = 1000000000


##stormDataSummarised <- select(stormData, EVTYPE, FATALITIES, INJURIES) %>% mutate(fatalitiesAndInjuriesCombined = FATALITIES + INJURIES) %>% group_by(EVTYPE) %>% summarize(countFatalities = sum(FATALITIES), countInjuries = sum(INJURIES), countFatalitiesAndInjutiesCombined = sum(fatalitiesAndInjuriesCombined)) %>% arrange(desc(countFatalities))
```

## Results

